---
title: |-
  Ecological Data Tutorial --
  <br> Spatial Analysis 
author: "Rob Colautti"
output: html_document
---

<script src="_hidecode.js"></script>

# 1. Background


# 2. Data

## `rgbif`, a GBIF API
A good source of species occurrence data is The Global Biodiversity Information Facility (GBIF) -- [gbif.org](https://www.gbif.org/tool/81747/rgbif)

There is a related R-package called `rgbif`, which acts as an API for accessing data from GBIF using R (API = Application Programming Interface). Here are links to the instructions and tutorial

    * [rgbif main site](https://www.gbif.org/tool/81747/rgbif)
    * [rgbif tutorial](https://ropensci.org/tutorials/rgbif_tutorial/)

Take a look at the website to understand the data before continuing, and refer back to the website to understand the type and structure of data.

Install and load the rgbif package
```{r, eval=F}
install.packages("rgbif")
```
```{r, results='hide'}
library(rgbif)
```

Look for occurrence data on garlic mustard (**Alliaria petiolata**). Note for that many species there could be thousands to hundreds of thousands of records. This can take a while to download, so we'll put an upper limit of 200 records to keep it from running too long. We also include only records that contain lat/long data
```{r}
GMgbif<-occ_search(scientificName="Alliaria petiolata", hasCoordinate=T, limit=200)
```

The function `occ_search()` returns a list. It's worth exploring the different items.
```{r}
names(GMgbif)
```

We're most interested in the `GMgbif$data` layer, which includes lats/longs. You can see the names by using the `names()` function on the `$data` slice:
```{r}
names(GMgbif$data)
```

## Finalize data

Extract Lats/longs into a new data.frame object, and rename columns
```{r}
gbifLOC<-as.data.frame(GMgbif$data[,grep("decimal",names(GMgbif$data))])
names(gbifLOC)<-c("Lat","Long")
head(gbifLOC)
```

# 3. `ggplot2` maps

You can create quick, rudimentary maps in ggplot2 using `borders()`
```{r}
library(ggplot2)
MyMap <- ggplot() + borders("world", colour="grey70", fill="black") + theme_bw()
MyMap
```

Add locations
```{r}
MyMap <- MyMap + geom_point(aes(x=gbifLOC$Long,y=gbifLOC$Lat),colour="green",size=2,alpha=0.1)
MyMap
```

Let's try to zoom in to the area covered by the points.
```{r}
MyMap + xlim(-100,50) + ylim(25,75)
```

What's wrong with this map? The borders are vectors that connect points are outside of the limits we set. When we set `xlim()` + `ylim()` ggplot cuts all those points, forcing redrawn polygons that no longer correspond with the borders. Look at `?borders` to find a solution. Note that there is xlim() and ylim() within the `borders()` function.
```{r}
MyMap <- ggplot() + borders("world", colour=NA, fill="black",
                            xlim=c(-100,50), ylim=c(30,75)) +
  geom_point(aes(x=gbifLOC$Long,y=gbifLOC$Lat), colour="green",
             size=2, alpha=0.1) + theme_bw() 
MyMap 
```

Now the map borders are complete, but to draw them `borders()` expands the x- and y-limits beyond what we defined so that all the relevant points are included. For example, our right xlim is 50 degrees east, but the plot area extends > 200 to plot the Russian border.

Another solution is `coord_fixed()`.
```{r}
MyMap + coord_fixed(xlim=c(-100,50), ylim=c(30,75))
```

Another problem is that we have 'hard coded' the x- and y-limits. What if we have a big data project and we want to have the borders adjust based on the data? Edit the above code to do that.

<div class="fold s o">
```{r}
# Add/subtract 1 degree to add a small buffer around the data points
PopRange<-c(min(gbifLOC$Long,na.rm=T)-1,
         min(gbifLOC$Lat,na.rm=T)-1,
         max(gbifLOC$Long,na.rm=T)+1,
         max(gbifLOC$Lat,na.rm=T)+1)
MyMap + coord_fixed(xlim=PopRange[c(1,3)], ylim=PopRange[c(2,4)])
```
</div>

## Datum/Projections

Do any of the points in the above graphs look odd (e.g. in the Ocean)? When working with geographic data you have to pay close attention to the __projection__ and __datum__ used. The problem is that lats/longs are based on a spheroid that have to be 'projected' to a flat surface for plotting. The datum describes the spheroid surface, since the earth is not a perfect sphere. The projection transforms the locations from a curved surface to a flat map. The datum you have used probably used is probably WGS84, which is the one used by Google Maps.

The datum information for `gbif` is available in the same `data.frame()` that we got lat/long info from:
```{r}
unique(GMgbif$data[46])
```

The misplacement of points is probably due to different projections for the mapping layer and the gbif point layer. Take a look at `?coord_map` for info on using projections.

```{r}
MyMap <- ggplot() + borders("world", colour=NA, fill="black") +
  coord_map(projection="cylindrical") + coord_fixed(xlim=PopRange[c(1,3)], ylim=PopRange[c(2,4)])
MyMap + geom_point(aes(x=gbifLOC$Long,y=gbifLOC$Lat), colour="green",
             size=2, alpha=0.1) + theme_bw()
```

If this didn't fix the problem, you might need a more complicated adjustment using projections in the `rgdal` package. We'll come back to this in the next section on ggmap. 

# 3. `ggmap` maps 

The `ggmap` package in R has nice functions for creating maps and plotting spatial data using online map APIs (e.g. Google Maps). 
```{r, eval=F}
install.packages("ggmap")
```
```{r, results='hide'}
library(ggmap)
```

The functions of `ggmap` are similar to `ggplot2`, so you might want to brush up on the tutorials for:

    1. [qplot visualizations](../RCrashCourse/2_qplot.html)
    2. [ggplot visualizations](../RCrashCourse/2_qplot.html)
    
Then check out this wonderful [ggmap cheat sheet](./ggmapCheatsheet.pdf) by Melanie Frazier

## Base map

The `ggmap` function is the plotting equivalent of `ggplot`. However, a lot of the action happens with `ggmap`. There are a lot of nice options, which you can view in the help file `?ggmap`

To map the base layer, we have to define a range of lats/longs. We'll define these values directly from the data (i.e. min/max of each lat/long with 1-degree buffer)

We define a mapping area as an address 
```{r}
ggmap(get_map(location="QUBS",source="google",maptype="satellite",zoom=14))
```

or lat/long midpoint __Here's a tribute to Gord Downie: `Where the Great Plains begin!`__
```{r}
ggmap(get_map(location=c(-100,49),source="google",maptype="satellite",zoom=8))
```

If you look at `?ggmap` you can see there is also an option for a bounding box: c(left-Long, bottom-Lat, right-Long, top-Lat). Rather than hard-code, let's use the data to define it:
```{r}
PopRange<-c(min(gbifLOC$Long,na.rm=T)-1,
         min(gbifLOC$Lat,na.rm=T)-1,
         max(gbifLOC$Long,na.rm=T)+1,
         max(gbifLOC$Lat,na.rm=T)+1)
ggmap(get_map(location=PopRange,source="google",maptype="road"))
```

Currently not working for Google Maps. Try another map source:
```{r, error=TRUE}
ggmap(get_map(location=PopRange,source="osm",maptype="satellite"))
ggmap(get_map(location=PopRange,source="stamen",maptype="satellite"))
ggmap(get_map(location=PopRange,source="cloudmade",maptype="satellite"))
```

Nope. Let's use our intuition from ggplot() to try to jury-rig Google Maps
```{r}
### First create a map object as above, zooming out far enough to cover the area
MyMap <- get_map(location=PopRange,source="google",maptype="satellite",zoom=3) 
### Then set Long & Lat limits
ggmap(MyMap) + coord_fixed(xlim=PopRange[c(1,3)], ylim=PopRange[c(2,4)]) 
```

Looks much better. Now let's add our point data:
```{r}
### First create a map object as above, zooming out far enough to cover the area
MyMap <- get_map(location=PopRange,source="google",maptype="satellite",zoom=3) 
### Then set Long & Lat limits
ggmap(MyMap) + coord_fixed(xlim=PopRange[c(1,3)], ylim=PopRange[c(2,4)]) +
  geom_point(aes(x=Long,y=Lat), data=gbifLOC, colour="yellow",
             size=2, alpha=0.1) 
```

Now another problem. Data points are showing up where they shouldn't (e.g. ocean). One common reason for this is that the map and the data points have different projections. We can use the `rgdal` package to fix this.

First take a look at the type of data we are using:
```{r}
class(gbifLOC)
```

We have to project the plotting data using the same datum and projection as the background map. For Google Earth Maps the datum is __WGS84__ and the projection is __Cylindrical__. This requires additional information attached to our simple lat/long dataset. The package `rgdal` has tools for this.
```{r, eval=F}
install.packages("rgdal")
```
```{r}
library(rgdal)
```

```{r}
class(gbifLOC)
```

Define the structure of the spatial data, using the names of the columns in the `data.frame()` object.
```{r}
gbifLOCsp<-gbifLOC
coordinates(gbifLOCsp)<-~Long+Lat
class(gbifLOCsp)
```

Assign the projection info using `proj4string()`. For more information on the CRS function, see this [helpful overview](./CRSoverview) also by Melanie Frazier
```{r}
proj4string(gbifLOCsp)<-CRS("+init=epsg:4326")
proj4string(gbifLOCsp)
```

Now we can project the lats/longs of the locations back to the map. But what projection to use?
```{r}
names(ggmap(MyMap))
```
Let's take a look at the coordinates layer.
```{r}
ggmap(MyMap)$coordinates
```

Under 'projection' it shows __mercator__. Probably the Web Mercator projection, which you can read about on [Wikipedia](https://en.wikipedia.org/wiki/Web_Mercator).

For a more comprehensive database, see [spatialreference.org](http://spatialreference.org) and [Proj.4](http://proj4.org)

Here is the Web Mercator projection and a few others.
```{r}
WebMerc<-CRS("+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m +nadgrids=@null +wktext  +no_defs")
Google<-CRS("+init=epsg:4326 +proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") # From Frazier's CRS document, linked above
Earth<-CRS("+proj=longlat +init=epsg:4326") # Alternate Google Earth Projection
Geo <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")# Projection from: http://pakillo.github.io/R-GIS-tutorial/
GMaps<-CRS("+proj=longlat +init=epsg:3857 +datum=WGS84") # Google Maps Projection
GBIF<-CRS("+init=epsg:4326")
```

```{r}
gbifLOCtform<-spTransform(gbifLOCsp,WebMerc)
```

Now that we have projected the datapoints, we can try plotting.

But FIRST, we have to transform back into a `data.frame()` object for the `ggmap()` function.
```{r}
gbifLOCtform_data<-data.frame(gbifLOCtform)
```

```{r}
ggmap(MyMap) + geom_point(aes(x=Long,y=Lat), data=gbifLOCtform_data, colour="yellow", size=2, alpha=0.1) 
```

Everything looks good, now let's rescale the map by recalculating the limits from the projected data.
```{r}
PopRange_tform<-c(min(gbifLOCtform_data$Long,na.rm=T)-1,
         min(gbifLOCtform_data$Lat,na.rm=T)-1,
         max(gbifLOCtform_data$Long,na.rm=T)+1,
         max(gbifLOCtform_data$Lat,na.rm=T)+1)
ggmap(MyMap)  +
  geom_point(aes(x=Long,y=Lat), data=gbifLOCtform_data, colour="yellow",
             size=2, alpha=0.1) + coord_fixed(xlim=PopRange_tform[c(1,3)], ylim=PopRange_tform[c(2,4)])
```

There is a problem when we rescale the map. Can you figure out what is happening?

The points and the background map are not scaling properly. 
```{r}
Rat<-abs((PopRange_trans[2]-PopRange_trans[4])/(PopRange_trans[1]-PopRange_trans[3]))
ggmap(MyMap)  +  geom_point(aes(x=Long,y=Lat), data=gbifLOCtform_data, colour="yellow",size=2, alpha=0.1) + coord_fixed(ratio=Rat)
```

Points are in the right spot, but now the map is stretched. There is no elegant solution here. As far as I know, the best we can do is hack the `get_googlemap()` function and play around with the `size=` and `scale=` parameters
```{r}
MidPoint<-c((PopRange_trans[1]+PopRange_trans[3])/2,(PopRange_trans[2]+PopRange_trans[4])/2)
MyMap<-get_googlemap(center=c(MidPoint),size=c(640,ceiling(640*Rat)),scale=2,source="google",maptype="satellite",zoom=2) 
ggmap(MyMap) +  geom_point(aes(x=Long,y=Lat), data=gbifLOCtform_data, colour="yellow",size=2, alpha=0.1) 
```




# 4. Extract environmental variables (simple)

# 5. Extract weather data (hard)

# 6. Test for spatial autocorrelation




