---
title: "Metabarcoding Analysis"
output: html_document
---

# Setup

```{r Load Packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
library(RAM)
library(lmtest)
library(car)
library(coefplot)
```

## data

Create a directory called `Data` inside of your working directory. Then copy these files into your Data folder:

  * [BioDatFull2.csv](./Data/BioDatFull2.csv)
  * [mapping_file.csv](./Data/mapping_file.csv)
  * [97_OTU_table_QIIME1.txt](./Data/97_OTU_table_QIIME1.txt)

```{r Loading data}
BioDatFull <- read.csv("./Data/BioDatFull2.csv")
Mapping_file <- read.csv("./Data/mapping_file.csv",fileEncoding="UTF-8-BOM")
OTU_table <- read.delim("./Data/97_OTU_table_QIIME1.txt", header=T)
names(Mapping_file)
```


We need to rearrange the OTU table. Think of this table exactly as you did the floristic survey! 

We want the species names at the top, and the sites are analogous to the samples (SRR...).

```{r}
row.names(OTU_table)<-OTU_table$OTU_ID
OTU_table$OTU_ID<-NULL

OTU_table.nz<-OTU_table[rowSums(OTU_table[,c(1:(length(OTU_table)-1))])!=0,]
OTU_table.ns<-OTU_table.nz[rowSums(OTU_table.nz[,c(1:(length(OTU_table.nz)-1))])!=1,] # We want to remove singletons because they are likely to be errors, especially in such a large dataset. 
dim(OTU_table)
dim(OTU_table.ns)

OTU_table<-OTU_table.ns

OTU_table.t<-transpose.OTU(OTU_table) 

which( colSums(OTU_table.t)==0 )
which( rowSums(OTU_table.t)==0 )
```


# Analysis

## Species alpha diversity model

### Calculate


```{r}
OTUmapping<-data.frame(SampleID=rownames(OTU_table.t))
row.names(OTUmapping)<-rownames(OTU_table.t)

OTUmapping$Richness<-rowSums(OTU_table.t!=0)
OTUmapping$Shannon<-diversity(OTU_table.t, index = "shannon", MARGIN = 1, base = exp(1))

Mapping_file$Species<-as.character(Mapping_file$Species) 
Mapping_file$SampleID<-as.character(Mapping_file$SampleID) 
OTUmapping$SampleID<-as.character(OTUmapping$SampleID) 
for(i in 1:(length(row.names(OTUmapping)))){
OTUmapping$Species[i]<-Mapping_file[Mapping_file$SampleID==OTUmapping$SampleID[i],]$Species
}
```

Unlike before, the "site" information is not attached to the file, its in a seperate CSV. 

We have two types of information: climatic info for each species (BioDatFull), and then species for each sample (Mapping_file). Now we also have our Richness and diversity information (OTU.info). We need to merge all of this and make just one dataset with the richness per species and the location/ climatic information.

```{r}
OTUmapping %>% group_by(Species) %>% dplyr::summarise(mean.Richness=mean(Richness),mean.Shannon=mean(Shannon)) -> OTU.info
OTU.info<-merge(OTU.info, BioDatFull, by="Species")
OTU.info$Species<-as.factor(OTU.info$Species)
MyData<-OTU.info
```

Now we model this. 

STEP 1. Fit the response data to a known distribution

```{r}
mod1<-lm(mean.Richness~ Lat + Long + Annual.mean.temperature.C + Max.temperature.C+ Min.temperature.C + Temperature.annual.range.C+ Annual.precipitation.mm. + Precipitation.seasonality.C.of.V.+ Annual.mean.moisture.index + Highest.weekly.moisture.index+ Lowest.weekly.moisture.index+ Moisture.index.seasonality.C.of.V., data = MyData)

car::qqp(MyData$mean.Richness, "norm")
plot(density(MyData$mean.Richness)) # does it fit a normal distribution?
shapiro.test(MyData$mean.Richness) 
bptest(mod1) # are the residuals homoskedastic?
plot(mod1) # visual check
```

STEP 2. Likelihood ratio test (test parameters)

To test each parameter, remove it and see if the model is as 'likely' without it. 


```{r}
mod1.mLat<-update(mod1,.~. - Lat)
anova(mod1,mod1.mLat)

mod1.mLong<-update(mod1,.~. - Long)
anova(mod1,mod1.mLong)

mod1.mAnnual.mean.temperature.C<-update(mod1,.~. - Annual.mean.temperature.C)
anova(mod1,mod1.mAnnual.mean.temperature.C)

mod1.mMin.temperature.C<-update(mod1,.~. - Min.temperature.C)
anova(mod1,mod1.mMin.temperature.C)

mod1.mTemperature.annual.range.C<-update(mod1.mMin.temperature.C,.~. - Temperature.annual.range.C)
anova(mod1.mMin.temperature.C,mod1.mTemperature.annual.range.C)

mod1.mAnnual.precipitation.mm.<-update(mod1.mMin.temperature.C,.~. - Annual.precipitation.mm.)
anova(mod1.mMin.temperature.C,mod1.mAnnual.precipitation.mm.)

mod1.mPrecipitation.seasonality.C.of.V.<-update(mod1.mMin.temperature.C,.~. - Precipitation.seasonality.C.of.V.)
anova(mod1.mMin.temperature.C,mod1.mPrecipitation.seasonality.C.of.V.)

mod1.mAnnual.mean.moisture.index<-update(mod1.mMin.temperature.C,.~. - Annual.mean.moisture.index)
anova(mod1.mMin.temperature.C,mod1.mAnnual.mean.moisture.index)

mod1.mHighest.weekly.moisture.index<-update(mod1.mAnnual.mean.moisture.index,.~. - Highest.weekly.moisture.index)
anova(mod1.mAnnual.mean.moisture.index,mod1.mHighest.weekly.moisture.index)

mod1.mLowest.weekly.moisture.index<-update(mod1.mAnnual.mean.moisture.index,.~. - Lowest.weekly.moisture.index)
anova(mod1.mAnnual.mean.moisture.index,mod1.mLowest.weekly.moisture.index)

mod1.mMoisture.index.seasonality.C.of.V<-update(mod1.mAnnual.mean.moisture.index,.~. - Moisture.index.seasonality.C.of.V.)
anova(mod1.mAnnual.mean.moisture.index,mod1.mMoisture.index.seasonality.C.of.V)

mod1<-mod1.mAnnual.mean.moisture.index
summary(mod1) # this is not done to look at the "stars" but rather to look at the Estimate value.
```

### Visualize

```{r}
coefplot(mod1)


avPlots(mod1)
```

## Redundancy analysis: a multivariate linear model

### Calculate

We have the same problem here with the samples having no specific data but rather we have experiement specific characteristics. 
We need to average them out then. 

```{r, message=FALSE, warning=FALSE}
X<-OTUmapping[,c(1,4)]
OTU_table.t$SampleID<-row.names(OTU_table.t)
OTU_table.t_species<-merge(OTU_table.t,X, by="SampleID",all = T)
OTU_table.t_species$SampleID<-NULL


OTU_table.tX<-aggregate(OTU_table.t_species, by = list(Species=OTU_table.t_species$Species), FUN=mean)
rownames(OTU_table.tX)<-OTU_table.tX$Species
OTU_table.tX$Species<-NULL
OTU_table.tX$Species<-NULL
OTU_table.hell <- decostand(OTU_table.tX, "hell")
```

Now we do the RDA:

```{r}
var <-MyData[,c(4:15)] # Isolate my variables to make this whole code easier to read

formulaRDA<- rda(OTU_table.hell ~ ., data=var, scale=T) # notice my predictive variable are being scaled too.

head(summary(formulaRDA))
RsquareAdj(formulaRDA)
anova(formulaRDA, permutations=1000) # Is the RDA meaningful?
anova.cca(formulaRDA, by="margin", permutations=1000) # If so, are any predictor variables meaningful?

```


### Visualize

Keep in mind none of these are significant though! so their direction does not have a true effect on the species composition. 


```{r}
smry <- summary(formulaRDA)
df1  <- data.frame(smry$sites[,1:2])       # RDA1 and RDA2 (for each quadrate)
df2  <- data.frame(smry$biplot[,1:2])     # loadings for RDA1 and RDA2 (for each variable)
rda.plot <- ggplot(df1, aes(x=RDA1, y=RDA2)) + 
  geom_text(aes(label=rownames(df1)),size=3,position=position_jitter(width=.2,height=.2)) +
  geom_point(aes(alpha=0.3)) +
  geom_hline(yintercept=0, linetype="dotted") +
  geom_vline(xintercept=0, linetype="dotted")  


(formulaRDA.PLOT<-rda.plot +
  geom_segment(data=df2, aes(x=0, xend=RDA1, y=0, yend=RDA2), 
               color="red", arrow=arrow(length=unit(0.01,"npc"))) +
  geom_text(data=df2, 
            aes(x=RDA1,y=RDA2,label=rownames(df2),
                hjust=0.5*(1-sign(RDA1)),vjust=0.5*(1-sign(RDA2))), 
            color="red", size=4)+ theme(legend.position="none"))
```


## NMDS: unconstrained spatial analysis

So this is when the NMDS can be really helpful. You dont know what is playing a role in the OTU community assembly. Looking at them in space might give you a hint, or might inspire you. 

### Calculate

We need to produce a betadiversity matrix again. We can keep the samples seperate now since we arent using climate here. We might be able to see on this if something else is happening. 

```{r}
df<-data.frame(Species=OTU_table.t_species$Species)

M<-OTU_table.t
M$SampleID<-NULL
M$Species<-NULL
M$Species<-NULL


dist_M <- vegdist(M, method = "bray",binary=T)
```


```{r}
meta.nmds <- metaMDS(dist_M, k=2, trymax = 1000) # ok so k=2 means I want 2 dimensions 


#data for plotting 
##NMDS points
NMDS.data<-data.frame(Species=df$Species) # this is the only true difference from floristic composition analysis. Instead of using plots location and population, we want to visualize their species. 
NMDS.data$NMDS1<-meta.nmds$points[ ,1] 
NMDS.data$NMDS2<-meta.nmds$points[ ,2] 
```


### Visualize

```{r}
ggplot(data = NMDS.data, aes(y = NMDS2, x = NMDS1))+ 
    geom_point( aes(col = NMDS.data$Species), size = 1.5,alpha=0.6)
```

They are not assembling because of climate... so what do you think might play into their assembly? This NMDS can help you think about it... 


