---
title: "Root Colonization Model"
author: "Robert I. Colautti"
date: "March 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Model

## Joint Probabilities

Colonization of roots by _k_ taxa at each of _N_ root intersections can be modelled as the joint probability of _k_ Bernoulli variables. A simple example is the joint probability of mycorrhizal colonization ($P_1$) and pathogen colonization ($P_2$). For each pair of taxa, the joint probability is:

$$ P(P_1=n_1 \space , \space P_2=n_2) $$

$$= P(P_2=n_2 | P_1=n_1) \cdot P(P_1=n_m)$$ 

$$= P(P_1=n_1 | P_2=n_2) \cdot P(P_2=n_2)) $$

where n~m~ and n~p~ are the binary colonization outcomes (0 or 1). 

Now consider the joint probabilities of each of the four possible outcomes:

$$A. \space Both := P(P_1 = 1 , P_2 = 1)$$

$$B. \space Mycorrhizae \space Only := P(P_1 = 1 , P_2 = 0)$$

$$C. \space Pathogens \space Only := P(P_1 = 0 , P_2 = 1)$$ 

$$D. \space Neither := P(P_1 = 0, P_2 = 0)$$

The variance-covariance matrix for mycorrhizae (1) and pathogens (2) species is:

$$V_m = (A + B)(1 - (A + B))$$

$$V_p = (A + C)(1 - (A + C))$$

$$Cov_{mp} = A - (A + B)(A + C)$$

The correlation coefficient between species is:

$$\frac {Cov_{mp}}{\sqrt{V_m \cdot V_p}} $$

$$= \frac {A - (A + B)(A + C)}{\sqrt {(A + B)(1 - (A + B))(A + C)(1 - (A + C))}}$$

## Null Model

Let $p_1$ be the colonization probability for Mycorrhizae

Let $p_2$ be the colonization probability for Pathogens

If colonzation probabilities are independent, then the joint probabilities are:

$$ A. \space Both = p_1p_2 $$

$$ B. \space MycOnly = p_1(1-p_2) $$

$$ C. \space PathOnly = (1-p_1)p_2$$

$$ D. \space Neither = (1-p_1)(1-p_2) $$

## Symmetric interference model

Now consider a symmetric interference model, where species interfere and exclude colonization by the other with probability $\alpha$. The joint probabilities are then:

$$ A. \space Both = p_1p_2 (1-2*\alpha)  $$

$$ B. \space MycOnly = p_1(1-p_2)+p_1P2*\alpha $$

$$ C. \space PathOnly = (1-p_1)p_2+p_1P2*\alpha$$

$$ D. \space Neither = (1-p_1)(1-p_2) $$

## Statistical Model

Given that $A$, $B$, $C$ & $D$ can be estimated from the data, it is easy to use the __Null Model__ to derive an equation for estimating parameters $p_1$ and $p_2$ from the observed data.



### The Challenge

Given that $A$, $B$, $C$ & $D$ can be estimated from the data, use the __Symmetric Interference Model__ to solve for $\alpha$, $p_1$ and $p_2$.

```{r}
Icor<-function(p1,p2,a){
  # Conditional probabilities
  ## A through B can be estimated from the data
  ## Use these equations to try to solve for p1, p2 & a
  ## For example, B-C or C-B will isolate p1 and p2 only
  A<-p1*p2*(1-2*a)
  B<-p1*(1-p2)+p1*p2*a
  C<-(1-p1)*p2+p1*p2*a
  D<-(1-p1)*(1-p2)

  # Equations for calculating (co)variances and correlation coefficient  
  V1<-(A+B)*(1-(A+B))
  V2<-(A+C)*(1-(A+C))
  COV12<-A-(A+B)*(A+C)

  COR12<-COV12/(sqrt(V1)*sqrt(V2))
  
  # Insert equations for A through B here
  ## For example, if you wanted to solve for p1 & p2 only
  p1_p2<-B-C
  ## Combine A through B to isolate each of these terms
  a<-NA # Insert equation for estimating alpha (a)
  p1<-NA # Insert equation for p1
  p2<-NA# insert equation for p2
  
  return(cbind(a,p1,p2))
}

# Test your model
SimDat<-data.frame(P1=runif(5000),P2=runif(5000),a=runif(5000))
Parm<-Icor(p1=SimDat$P1,p2=SimDat$P2,a=SimDat$a)
library(ggplot2)

#Each of these 3 plots will be a 1:1 line if your calculations are correct"
## Test p1
qplot(SimDat$P1,Parm[,2])
## Test p2
qplot(SimDat$P2,Parm[,3])
## Test alpha
qplot(SimDat$P2,Parm[,1])
```

